---
title: "MiDAS PhD course 2024"
author: "Marie Riisgaard-Jensen and Sofie Zacho Vestergaard"
date: "2024-05-22"
output: html_document
---

# Installation of packages (!!only do once!!)
```{r}
#ampvis2
install.packages("remotes")
remotes::install_github("kasperskytte/ampvis2")

#dplyr for data manipulation
install.packages("dplyr")
```


# Load packages
```{r, include=FALSE}
library(ampvis2)
library(dplyr)
```


# Load data
```{r}
# Set paths
V13_metadata = 'Amplicon_data_for_hands_on_exercices/V13_metadata.txt'#metadata
V13_ASVtab = 'Amplicon_data_for_hands_on_exercices/V13_ASVtab.txt' #ASV table
V13_tax = 'Amplicon_data_for_hands_on_exercices/V13_ASV_MiDAS_5.3.sintax' #taxonomy
ASV='Amplicon_data_for_hands_on_exercices/V13_ASV.fa' #ASVs

# Load into ampvis object
d13 <- amp_load(otutable = V13_ASVtab, 
                taxonomy = V13_tax,
                metadata = V13_metadata,
                fasta = ASV)

# Add location to metadata (based on GPS coordinates) and tank ID
d13$metadata <- d13$metadata %>% 
  mutate(Location = paste0(Latitude, ",",Longitude),
         Tank_ID = substr(SampleID, 1, nchar(SampleID) - 2))

```


# Explore metadata
```{r}
# extract metadata from ampvis object
meta <- d13$metadata

print(meta)

```

```{r}
# number of cities
unique(meta$City)
```

```{r}
# How many samples from each city
table(meta$City)
```

```{r}
table(meta$Location)
```

```{r}

# Convert Location to a factor with the new order (only used for plotting)
meta$Location <- factor(meta$Location, levels = unique(meta$Location[order(meta$City)]))

# plot number of samples for each plant
ggplot(meta,aes(x=Location, fill = City)) +
  geom_bar(width = 0.8) +
  labs(y= 'Number of samples')+
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_brewer(palette = 'BrBG')
  

```

```{r}
# put on numbers of tanks for each plant
meta %>% group_by(Location) %>% 
  mutate(Tanks = n_distinct(Tank_ID)) %>%
  ggplot(.,aes(x=Location)) +
  geom_bar(aes(fill = City),width = 0.8) +
  labs(y= 'Number of samples')+
  geom_label(aes(y=Tanks,label=Tanks)) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_brewer(palette = 'BrBG')
  
```

```{r}
install.packages('ggrepel')
library(ggrepel)

metadata <- meta
world_coordinates <- ggplot2::map_data("world") %>%
  full_join(., metadata %>% rename("region" = "Country"),
            relationship = "many-to-many", by = join_by(region)) %>%
  ungroup() %>%
  filter(region == "Australia")


  ggplot() +
  geom_map(
    data = world_coordinates, map = world_coordinates,
    aes(long, lat, map_id = region,), 
    color = "grey", linewidth = 0.1, fill = "grey90",
  ) +
  geom_point(data = distinct(metadata, City, Latitude, Longitude),
             aes(x = Longitude, y = Latitude, color = City),
             size = 3.4) +
  geom_label_repel(data = distinct(metadata, City, Latitude, Longitude),
             aes(x = Longitude, y = Latitude, label = City, fill = City),
             alpha = 0.8,
             size = 3.4) +
  theme_void() +
    theme(legend.position = "none")
```



## Explore sequences
```{r}
d13$refseq
```







# Number of reads

## What are the minimum and maximum number of reads? 
```{r}
d13
```

## What are the minimum and maximum number of reads of city13?
```{r}
d13_city13 <- amp_subset_samples(d13, City == 'City13')
d13_city13
```

## Have the samples been sequenced deep enough?
```{r}
amp_rarecurve(d13,
              facet_by = 'City', color_by = 'Location')
```

## Rarefy the data 
```{r}

d13_rare <- amp_subset_samples(d13, rarefy = 10000, minreads = 10000, 
                               removeAbsentOTUs = T)

```


# Alpha diversity of data
## Calculate the alpha diversity (ObservedOTU, Shannon, Inverse Simpson)
```{r}
alpha <- amp_alpha_diversity(d13_rare)

ggplot(alpha, aes(x = Location,
                  y = Shannon,
                  color = City)) +
  geom_point()

```










# Normalise the data
## 
```{r}
d13n <- amp_subset_samples(d13_rare,normalise = T)

d13_rare$abund
d13n$abund
```




# Most abundant bacteria
## Which bacterial taxa are the most abundant across all samples?
```{r, fig.height=2, fig.width=6}
amp_heatmap(d13n,
            normalise = F,
            tax_aggregate = "Genus",
            tax_add = "Phylum",
            group_by = "Location",
            sort_by = "-27.55,152.99",
            tax_show = 20,
            showRemainingTaxa = T,
            plot_functions = T
            )
```





```{r}
amp_boxplot(d13n,
            group_by = 'City')


```






# Beta diversity
## Make a ordination plot of all samples
```{r}
amp_ordinate()

```


